\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-04-13 honors_thesis/journal_paper/paper.tex
Some fixes to Andrew's fixes; pretty close to done, I think

Changing the
neighborhood topology has a significant effect on the performance of the
algorithm and this effect varies widely by function type.  The two most common
topologies used in the literature are the Ring topology and the Complete
topology~\cite{bratton-2007-defining-a-standard-for-pso}.  In the Ring
topology, each particle has one neighbor to either side of it; in the Complete
topology, every particle is a neighbor to every other particle.  In all
topologies a particle is also a neighbor to itself in that its own position and
value are considered when updating the particle's neighborhood best, $\nbest$.
Thus with $p$ particles, using the Ring topology each particle with index $i$
has three neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete
topology, each particle has $p$ neighbors.

@@

The PSO algorithm can be trivially parallelized by distributing the processing
needed for each particle on across processors.  But as we have seen in
\fig{fig:evals-sphere}, for some functions, and for large numbers of
processors, just adding particles reaches a point of diminishing returns.
That is, adding processors does not help us reach any given level of fitness
appreciably faster.  Instead of adding particles we employ a speculative
evaluation approach that allows us to perform iterations two at a time.

@@

The refactoring hinges on the idea that once the random coefficients are
determined, there are only a few possibilities for the new values of updates
to $\nbest$ and $\pbest$.  For the Ring topology there are 7 possible update
cases, identified in Table~\ref{tab:evals}.  We label each case with an
identifier referring to the source of the update: a minus sign ($-$)
represents no update, $L$ represents an update to $\nbest$ coming from the
left neighbor, $R$ represents an update to $\nbest$ coming from the right
neighbor, and $S$ represents an update to either $\pbest$ or $\nbest$ coming
from the particle itself.  As an example, $\casexn$ refers to the case that
the particle finds a new personal best, but neither it nor its neighbors found
a position that updated its neighborhood best.  In the equations that follow,
we refer to an unspecified update case as $\casegen$, and to the set of cases
collectively as $\caseset$.

@@

For each case $\casegen \in \caseset$, we introduce an indicator function
$\indic_{t}^{\casegen}$.  This function is 1 for the realized case from
Table~\ref{tab:evals} and 0 for the other cases.  We can then sum over all of
the cases, and the indicator function will make all of the terms drop to zero
except for the case that actually occurs.  Although this update is usually
implemented using if-statements, it is convenient to describe it as a
mathematical expression.  For example, the indicator function for the specific
case $\casexn$ can be written as follows:
\begin{align}
  \nonumber
	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1}))= \\
  \label{eq:deficasexn}
	&\begin{cases}
	   1 & \text{if} \ \ofunc(\ppos_{t}) < \ofunc(\pbest_{t-1}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\ppos_{t}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\leftn_{t}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\rightn_{t}) \\
	   0 & \text{otherwise}
	\end{cases}
\end{align}

@@

In the same way we can create notation for the position update function by
substituting into~\eqref{eq:posupdate}.  For compactness, we will drop the
parameters to $\specvel_{t+1}^{\casegen}$ since they can be inferred from the
subscripts on $\indic_{t+1}^{\casegen}$.
\begin{align}
\label{eq:defpcasegen}
	\specpos_{t+1}^{\casegen} & (\ppos_{t}, \pvel_{t}, \leftn_{t},
	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
\nonumber
	& = \ppos_{t} + \specvel_{t+1}^{\casegen}
\end{align}

>>>>

Changing the way neighbors are defined (often called the ``topology'' in the
literature) has a significant effect on the performance of the algorithm.  In
the Ring topology, each particle has one neighbor to either side of it; in the
Complete topology, every particle is a neighbor to every other
particle~\cite{bratton-2007-defining-a-standard-for-pso}.  In all topologies a
particle is also a neighbor to itself in that its own position and value are
considered when updating the particle's neighborhood best, $\nbest$.  Thus with
$p$ particles, using the Ring topology each particle with index $i$ has three
neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete topology, each
particle has $p$ neighbors.

@@

The PSO algorithm can be trivially parallelized by distributing the processing
needed for each particle on across processors.  But as we have seen in
\fig{evals-sphere}, for some functions, and for large numbers of
processors, just adding particles reaches a point of diminishing returns.
That is, adding processors does not help us reach any given level of fitness
appreciably faster.  Instead of adding particles we employ a speculative
evaluation approach that allows us to perform iterations two at a time.

@@

The refactoring hinges on the idea that once the random coefficients are
determined, there are only a few possibile updates to $\nbest$ and $\pbest$.
For the Ring topology there are 7 possible update cases, identified in
Table~\ref{tab:evals}.  We label each case with an identifier referring to the
source of the update: a minus sign ($-$) represents no update, $L$ represents
an update to $\nbest$ coming from the left neighbor, $R$ represents an update
to $\nbest$ coming from the right neighbor, and $S$ represents an update to
either $\pbest$ or $\nbest$ coming from the particle itself.  As an example,
$\casexn$ refers to the case that the particle finds a new personal best, but
neither it nor its neighbors found a position that updated its neighborhood
best.  In the equations that follow, we refer to an unspecified update case as
$\casegen$, and to the set of cases collectively as $\caseset$.

@@

In order to incorporate the determination of which case occurs into the
position and velocity update equations, we introduce an indicator function
$\indic_{t+1}^{\casegen}$ for each case $\casegen \in \caseset$.  When
$\casegen$ corresponds to the case actually taken by PSO,
$\indic_{t+1}^{\casegen}$ evaluates to 1; otherwise it evaluates to 0.  We can
then sum over all of the cases, and the indicator function will make all of the
terms drop to zero except for the case that actually occurs.  For example, the
indicator function for the specific case $\casexn$ can be written as follows:

\begin{align}
  \nonumber
	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1}))= \\
  \label{eq:deficasexn}
	&\begin{cases}
	   1 & \text{if} \ \ofunc(\ppos_{t}) < \ofunc(\pbest_{t-1}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\ppos_{t}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\leftn_{t}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\rightn_{t}) \\
	   0 & \text{otherwise}
	\end{cases}
\end{align}

@@

In the same way we can create notation for the position update function by
substituting into~\eqref{eq:posupdate}.  For compactness, we will drop the
parameters to $\specvel_{t+1}^{\casegen}$ since they can be inferred from the
subscripts.
\begin{align}
\label{eq:defpcasegen}
	\specpos_{t+1}^{\casegen} & (\ppos_{t}, \pvel_{t}, \leftn_{t},
	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
\nonumber
	& = \ppos_{t} + \specvel_{t+1}^{\casegen}
\end{align}
